### 2016-05-04 p01-ch02-2013/paper.tex
Fix ref/label for Chapter 02

\section{Introduction} \label{sec:introduction}

@@

The paper continues as follows: Section \ref{sec:deadlines} introduces the language constructs
and provides informal semantics of the language with a case study in Section \ref{sec:frh}. 
Section \ref{sec:sos} presents the operational semantics of Crisp.
Section \ref{sec:impl} follows to provide a detailed discussion on the implementation.
The case study continues in this section with further details and code examples.
Section \ref{sec:relwork} discusses related work of research and 
finally Section \ref{sec:conclusion} concludes the paper and proposes future line of research.

\section{Programming with deadlines} \label{sec:deadlines}
In this section, we introduce the basic concepts underlying  the notion of ``deadlines'' for asynchronous messages between actors. 
The main new constructs specify how a message can be sent with a deadline, how the message response can be processed, and what happens when a deadline is missed.
We discuss the informal semantics of these concepts  and illustrate them using a case study in Section \ref{sec:frh}.

Listing \ref{fig:grammar} introduces a minimal version of the  real-time
actor-based language Crisp.
Below we discuss the two main new language constructs presented at lines (7) and (8).
% In Line (6), we introduce an extension on how each concurrent object can send a message to another concurrent object annotated with an extra
% property that represents a deadline ($d$).

\begin{figure}[t]
\begin{align}
C &::= \textbf{\jtt{class}} \; \mathsf{N} \; \textbf{\jtt{begin}} \; V^? \; \{M\}^* \; \textbf{\jtt{end}} \\
M_{sig} &::=
%\textbf{\jtt{op}} \; 
\mathsf{N}(\overline{T \: x}) \\ 
M &::= \{ M_{sig} == \{V \; ;\}^? S \} \\
V &::= \textbf{\jtt{var}} \; \{\{x\},^+ \; : \; \mathsf{T} \; \{ = e\}^? \},^+ \\
S &::=  x := e\; | \\
 &::= x := \textbf{\jtt{new}} \; \mathsf{T}(e^?) \; |\\
&:= f = e \; ! \; m(\overline{e}) \; \textbf{\jtt{deadline}}(e) \: | \\
 &::= x := f.\gett(e^?) \; | \\
  &::= \textbf{\jtt{return}} \; e  \; | \\
  &::= S \; ; \; S \; | \\
  &::= \textbf{\jtt{if}} \; (b) \; \textbf{\jtt{then}} \; S \; \textbf{\jtt{else}} \; S \; \textbf{\jtt{end}}  \; | \\
  &::= \textbf{\jtt{while}} \; (b) \; \{ \; S \; \} \; | \\
  &::= \textbf{\jtt{try}} \: \{ S \} \: \jtt{\textbf{catch}}(\mathsf{T}_{\jtt{Exception}} \; x) \: \{ \: S \: \}  
%E &::= B \; | \; V \\
%B &::= \textbf{\jtt{true}} \; | \; \textbf{\jtt{false}} \; | \; B \wedge B \; | \; B \vee B \; | \; %\textbf{\jtt{not}} \; B
\end{align}
\caption[A kernel version of the real-time programming language]{
A kernel version of the real-time programming language.
The bold scripted \textbf{\jtt{keywords}} denote the reserved words in the language.
The over-lined $\overline{v}$ denotes a sequence  of syntactic entities $v$.
Both local and instance variables are denoted by $x$.
We assume distinguished local variables  $\jtt{this}$, $\jtt{myfuture}$, and  $\jtt{deadline}$ which
denote the actor itself, the  unique future corresponding to the process, and  its deadline,  respectively.
A distinguished instance variable $\jtt{time}$  denotes  the current time.
Any subscripted type $\mathsf{T}_{specialized}$ denotes a \textit{specialized} type of general type $\mathsf{T}$; 
e.g. $\mathsf{T}_{\jtt{Exception}}$ denotes all ``exception'' types.
A variable $f$ is in $\mathsf{T}_{future}$. 
\textsf{N} is a name (identifier) used for classes and method names.
%$e$ is in \textsf{Expression} ($E$);
$C$ denotes a class definition which consists of  a definition of its instance variables
and its methods;
$M_{sig}$ is a method signature; $M$ is a method definition; $S$ denotes a statement.
We abstract from the syntax the side-effect free expressions $e$ and
boolean expressions $b$.
%
%For simplicity, we drop the notation syntax to specify how $V$ can expand to different values along with arithmetic operations.
%\emph{Note} that small letters (e.g. $e$) are of type of their capital constructs (e.g. $E$).
}
\label{fig:grammar}
\end{figure}

@@

\parbf{What happens when a deadline is missed?}
Messages received by an actor generate processes.
Each actor contains one active process and all its other processes are queued. 
Newly generated processes are \emph{inserted} in the queue according to an application-specific policy.
When a \emph{queued} process misses its deadline it is \emph{removed} from the queue and a
corresponding \emph{exception} is recorded by its future (as described below).
When the currently active process is terminated the process at the head of the queue is activated
(and as such  dequeued).
The active process  cannot be \emph{preempted} and is forced
to  \emph{run to completion}.
In  Section \ref{sec:impl} we discuss the implementation details of this design choice.

@@

For example, in Listing \ref{lst:try:catch}, if the \gett \space operation raises an exception  control,  is transferred to line (5);
otherwise, the execution continues in line (3). 
In the \jtt{catch} block, the programmer has also access to the occurred exception that can be 
any kind of exception including an exception that is caused by a missed deadline.
In general, any uncaught exception gives rise to abortion of the active process and is recorded by its future.
Exceptions in our actor-based model thus are propagated by futures.

@@

% Since, we provide a reference implementation in Java (c.f. Section \ref{sec:impl}), we take advantage of exception ``stack traces''; 
% i.e. an instance of exception can produce a chain of ``causes'' that has led to the occurrence of the exception.
% This means that cathing an exception may in turn lead to throwing another exception or just rethrowing it. 
% Therefore, if the programmer decides not to use a \jtt{try-catch} block, the exception is propagated through the call hierarchy.
% It means that the propagation of the exception may also contribute to the stack trace of another exception.
% If any exception is not eventually caught, the execution environment will receive it and the application will stop.

% As discussed in detail in our previous work \cite{crisp-sac}, object $o'$ maintains a queue of processes for the incoming messages such as $m$.
% While a message $m$ is in the queue of $o'$, there is a chance that the deadline $d$ for message $m$ would be missed.
% We say a deadline $d$ for message $m$ on object $o'$ is missed when
% \begin{enumerate}
%  \item The process for message $m$ is \emph{not} the current process of $o'$. 
%  In other words, we do \emph{not} terminate the current process in case it missses its deadline; 
%  i.e. we allow \emph{run to completion} for the current process. And,
%  \item The response for message $m$ is not ready and the \emph{current} time has passed the deadline $d$ expectation.
% \end{enumerate}
% 
% When a deadline-miss is detected, it is required to be reflected in the future response that is being waited on in object $o$. 
% Therefore, for a deadline $d$ that is missed on message $m$:
% 
% \begin{enumerate}
%  \item The process that is related to message $m$ is aborted and removed from $o'$'s  process queue.
%  \item The aborted state is recorded in the corresponding future value $f$.
%  \item Whenever, \jtt{f.get()} is invoked, the \emph{aborted} state causes an exception. 
%  The exception is supposed to be handled by the user of the future; otherwise, it is propagated in the call hierarchy.
% \end{enumerate}
% 
% \emph{Starvation?} \Behrooz{Talk about the responsiveness of the system.}
% 
% \emph{Discussion?} Compare the approaches: event-based with no get() and exceptions as messages vs. messages with exceptions
\subsection{Case Study: Fredhopper Distributed Data Processing} \label{sec:frh}

\begin{figure}[h]
% \begin{wrapfigure}{r}{0.4\textwidth}
% \vspace{-40pt}
\begin{center}
\includegraphics[scale=0.4]{figs/DM}
\end{center}
% \vspace{-20pt}
\caption[Life cycle for remote data processing]{Fredhopper's Controller life cycle for remote data processing}
\label{fig:frh-dm}
% \vspace{-10pt}
% \end{wrapfigure}
\end{figure}

@@

To orchestrate different services such as FAS or data processing, Fredhopper takes advantage of a service controller (a.k.a. Controller).
Controller is responsible to passively manage different service installations for each customer.
For instance, in one scenario, a customer submits their data along with a processing request to their data hub server.
Controller, then picks up the data and initiates a data processing job (usually an ETL job) in a data processing service.
When the data processing is complete, the result is again published to customer environment and additionally becomes available through FAS services.
Figure \ref{fig:frh-dm} illustrates an example scenario that is described above.

@@

To illustrate the contribution of this paper, we extract a closed-world simplified version of the scenario in Figure \ref{fig:frh-dm} from Controller.
In Section \ref{sec:impl}, we provide an implementation-level usage of our work applied to this case study.

\section{Operational Semantics} \label{sec:sos}
We describe the semantics of the language by means of a two-tiered labeled transition system:
a local transition system describes the behavior of a single actor and a global transition system
describes the overall behavior of a system of interacting actors.
We define an actor state as a pair $\langle p,q \rangle$, where
\begin{itemize}
%  \item $o$ is the object identifier
 \item $p$ denotes the current active process of the actor, and
 \item $q$ denotes a queue of pending processes.
\end{itemize}
Each pending process is a pair  $(S, \tau)$ consisting of the current executing statement $S$ and the assignment $\tau$ of 
values  to the  \emph{local} variables (e.g., formal parameters).
The active process consists of a  pair  $(S,\sigma)$, where $\sigma$ assigns values to the local variables
and additionally assigns values to the instance variables
of the actor.

@@

\parbf{Incoming message.}
Let the active process $p$ belong to the actor $\tau(\jtt{this})$ (i.e., 
 $\sigma(\jtt{this})=\tau(\jtt{this})$ for the assignment $\sigma$ in $p$):
$$
\langle p , q, \phi\rangle \xrightarrow[]{m(\tau)}  \langle p , \mathsf{insert}(q, m(\overline{v}, d)), \phi\rangle
$$
where   $\mathsf{insert}(q, m(\tau))$  defines the result of inserting the process  $(S,\tau)$, where $S$ denotes the body of method $m$,
in $q$, according to some application-specific policy (described below in Section \ref{sec:impl}).

@@

% \subsection{Previously written in case notations would be reused}
% 
% \Behrooz{Previous wrong version. May be removed later.}
% \emph{Receive a new message.} The operational semantics when an object with identifier $o$ receives a message to invoke method $m$ with deadline $d$.
% 
% \begin{align}
% \frac
% {\mathsf{in} = m(\overline{v}, d) + \dot{p} = \mathsf{new()}}
% {(o, (\dot{p}.\jtt{deadline} \mapsto \mathsf{time()} + d,s[\overline{v}], d), p, \mathsf{enq}(q, \dot{p}), \sigma)}
% \end{align}
% 
% in which the operation \textsf{new()} creates a new instance of a process for the concurrent object.
% Additionally, as we discssued in Equation \ref{eq:orderedq}, each concurrent object owns an ordered queue for the processes.
% The process queue is ordered based on the result of $\sigma$ function:
% 
% $$ \sigma(\dot{p}) = \dot{p}.\jtt{deadline} $$
% 
% Thus, the operational semantics of \textsf{enq} ensures to find a proper index place $i$ for the new process $\dot{p}$ such that:
% 
% $$
% \forall k < i \::\: q_k.\jtt{deadline} < \dot{p}.\jtt{deadline} \wedge \forall k > i \::\: q_k.\jtt{deadline} > \dot{p}.\jtt{deadline}
% $$
% 
% And, then, $q_i = \dot{p}$ completes the enqueuing action. Since, $q$ is an \emph{ascending} ordered queue, the required action to deqeue the 
% ``next'' process is simply to remove and return the \emph{head} of the queue; namely $q_0$.
% 
% \emph{Deadline Miss Detection.} A deadline miss is detected by a separate watchdog thread of the active concurrent object.
% The watchdog thread observes all the processes but the current process of the object to allow it complete its execution:
% 
% \begin{align}
% \frac
% {\forall p \neq \overline{p} \::\: (o,s[f_p \mapsto \jtt{null}],\overline{p},q,\sigma) + p.\jtt{deadline} < \mathsf{time()}}
% {(o,s[f_p.\jtt{aborted} \mapsto \jtt{true}], \overline{p}, q - p, \sigma)}
% \end{align}
% 
% Consequently, when the future of a missed-deadline process $p$ is accessed, it causes the application to generate and throw an exception:
% 
% \begin{align}
% \frac
% {(o,(f.\jtt{get()},s[f.\jtt{aborted} = \jtt{true}],d),q,\sigma)}
% {(o,(x \mapsto \jtt{new Exception},s[f \mapsto \jtt{null}],d),q,\sigma)}
% \end{align}

\section{Implementation} \label{sec:impl}
We base our implementation on Java's concurrent package: \jtt{java.util.concurrent}. 
The implementation consists of the following major components:
\begin{enumerate}
 \item An extensible language API that owns the core abstractions, architecture, and implementation. 
For instance, the programmer may extend the concept of a scheduler to take full control of how, i.e., in what order, the processes of the individual actors are queued
(and as such scheduled for execution).
We illustrate the scheduler extensibility with an example in the case study below.
 \item Language Compiler that translates the modeling-level programs into Java source.
 We use ANTLR \cite{antlr} parser generator framework to compile modeling-level programs to actual implementation-level source code of Java.
 \item The language is seamlessly integrated with Java.
 At the time of programming, language abstractions such as data types and third-party libraries from either Crisp or Java are equally usable by the programmer.
\end{enumerate}

@@

\parbf{Fredhopper case study.}
As introduced in Section \ref{sec:frh}, we extract a closed-world simplified version from Fredhopper Controller.
We apply the approach discussed in this paper to use deadlines for asynchronous messages.
% Controller operates in a high-demand distributed setting as in Figure \ref{fig:ctl:deploy}.

Listing \ref{lst:frh:polling} and \ref{lst:frh:deadlines} present the difference in the previous Controller and the approach in Crisp.
The left code snippet shows the Controller that uses polling to retrieve data processing results.
The right code snippet shows the one that uses messages with deadlines.

@@

When the approach in Crisp in the right snippet is applied to Controller, it is guaranteed that all data job requests are terminated in a \emph{finite} amount of time.
Therefore, there cannot be complains about never receiving a response for a specific data job request.
Many of Fredhopper's customers rely on data jobs to eventually deliver an e-commerce service to their end users.
Thus, to provide a guarantee to them that their job result is always published to their environment is critical to them.
As shown in the code snippet, if the data job request is failed or aborted based on a deadline miss, 
the customer is still eventually informed about the situation and may further decide about it.
However, in the previous version, the customer may never be able to react to a data job request because its results are never published.
% \begin{wrapfigure}{r}{0.35\textwidth}
% \vspace{-35pt}
% \begin{center}
% \includegraphics[scale=0.35]{frh-ctl}
% \end{center}
% \vspace{-20pt}
% \caption{Fredhopper's Controller distribution model}
% \label{fig:ctl:deploy}
% \vspace{-15pt}
% \end{wrapfigure}

In comparison to the Controller using polling, there is a way to express timeouts for future values.
However, it does not provide language constructs to specify a deadline for a message that is sent to data processing service.
A deadline may be simulated using a combination of timeout and periodic polling approaches (Listing \ref{lst:frh:polling}).
Though, this approach cannot guarantee eventual termination in all cases; as discussed before that Step 4 in Figure \ref{fig:frh-dm} may never complete.
Controller is required to meet certain customer expectations based on an SLA.
Thus, Controller needs to take advantage of a language/library solution that can provide a higher level of abstraction for real-time scheduling of concurrent messages.
When messages in Crisp carry a deadline specification, Controller is able to guarantee that it can provide a response to the customer.
This termination guarantee is crucial to the business of the customer.

@@

In the above listings, Listing \ref{lst:frh:sched} defines a custom scheduler that determines the priority of two processes with custom logic for specific customer.
To use the custom scheduler, the only requirement is that the class \jtt{DataProcessor} defines a specific class variable called \jtt{scheduler} in Listing \ref{lst:frh:dm}.
The \textsl{custom scheduler} is picked up by Crisp core architecture and is used to schedule the queued processes.
% In the listing, \jtt{DefaultSchedulingManager} denotes the default EDF scheduler that is implemented in Crisp.
Thus, all processes from customer \jtt{A} have priority over processes from other customers no matter what their deadlines are.

@@

\begin{table}[h]
% \begin{wraptable}{l}{0.25\textwidth}
% \vspace{-15pt}
\begin{tabular}{c c}\thickhline
$s_1$ & $s_2$ \\ \thickhline
$88.71\%$ & $94.57\%$ \\ \hline
\end{tabular}
\caption{Evaluation Results}
\label{tbl:eval}
% \vspace{-15pt}
% \end{wraptable}
\end{table}

For a customer $c$, the satisfaction can be denoted by $s = \frac{r^F_c}{r_c}$; in which $r^F_c$ is the number of finished data processing jobs and $r_c$ is the total number of requested data processing jobs from customer $c$. 
We extracted statistics for completed and never-ended data processing jobs from Controller logs ($s_1$).
We replayed the logs with Crisp approach and measured the same property ($s_2$).
We measured the same property for 180 customers that Fredhopper manages on the cloud.
In this evaluation, a total number of about 25000 data processing requests were included.
The results show $6\%$ improvement in Table \ref{tbl:eval} (that amounts to around 1600 better data processing requests). 
Because of data issues or wrong parameters in the data processing requests, there are requests that still fail or never end and should be handled by a human resource.

@@

\section{Related Work} \label{sec:relwork}
The programming language presented in this paper is a real-time  extension of the  language introduced in  \cite{crisp-sac}. This new extension features
\begin{itemize}
 \item integration of  asynchronous messages with  deadlines and futures with timeouts;
 \item  a general mechanism for handling exceptions raised by missed deadlines;
  \item high-level specification of application-level scheduling policies; and
  \item a formal operational semantics.
\end{itemize}
To the best of our knowledge the resulting language is the first implemented 
real-time actor-based programming language which formally integrates the above features.

>>>>

\section{Introduction} \label{ch02:sec:introduction}

@@

The paper continues as follows: Section \ref{ch02:sec:deadlines} introduces the language constructs
and provides informal semantics of the language with a case study in Section \ref{ch02:sec:frh}. 
Section \ref{ch02:sec:sos} presents the operational semantics of Crisp.
Section \ref{ch02:sec:impl} follows to provide a detailed discussion on the implementation.
The case study continues in this section with further details and code examples.
Section \ref{ch02:sec:relwork} discusses related work of research and 
finally Section \ref{ch02:sec:conclusion} concludes the paper and proposes future line of research.

\section{Programming with deadlines} \label{ch02:sec:deadlines}
In this section, we introduce the basic concepts underlying  the notion of ``deadlines'' for asynchronous messages between actors. 
The main new constructs specify how a message can be sent with a deadline, how the message response can be processed, and what happens when a deadline is missed.
We discuss the informal semantics of these concepts  and illustrate them using a case study in Section \ref{ch02:sec:frh}.

Listing \ref{ch02:fig:grammar} introduces a minimal version of the  real-time
actor-based language Crisp.
Below we discuss the two main new language constructs presented at lines (7) and (8).
% In Line (6), we introduce an extension on how each concurrent object can send a message to another concurrent object annotated with an extra
% property that represents a deadline ($d$).

\begin{figure}[t]
\begin{align}
C &::= \textbf{\jtt{class}} \; \mathsf{N} \; \textbf{\jtt{begin}} \; V^? \; \{M\}^* \; \textbf{\jtt{end}} \\
M_{sig} &::=
%\textbf{\jtt{op}} \; 
\mathsf{N}(\overline{T \: x}) \\ 
M &::= \{ M_{sig} == \{V \; ;\}^? S \} \\
V &::= \textbf{\jtt{var}} \; \{\{x\},^+ \; : \; \mathsf{T} \; \{ = e\}^? \},^+ \\
S &::=  x := e\; | \\
 &::= x := \textbf{\jtt{new}} \; \mathsf{T}(e^?) \; |\\
&:= f = e \; ! \; m(\overline{e}) \; \textbf{\jtt{deadline}}(e) \: | \\
 &::= x := f.\gett(e^?) \; | \\
  &::= \textbf{\jtt{return}} \; e  \; | \\
  &::= S \; ; \; S \; | \\
  &::= \textbf{\jtt{if}} \; (b) \; \textbf{\jtt{then}} \; S \; \textbf{\jtt{else}} \; S \; \textbf{\jtt{end}}  \; | \\
  &::= \textbf{\jtt{while}} \; (b) \; \{ \; S \; \} \; | \\
  &::= \textbf{\jtt{try}} \: \{ S \} \: \jtt{\textbf{catch}}(\mathsf{T}_{\jtt{Exception}} \; x) \: \{ \: S \: \}  
%E &::= B \; | \; V \\
%B &::= \textbf{\jtt{true}} \; | \; \textbf{\jtt{false}} \; | \; B \wedge B \; | \; B \vee B \; | \; %\textbf{\jtt{not}} \; B
\end{align}
\caption[A kernel version of the real-time programming language]{
A kernel version of the real-time programming language.
The bold scripted \textbf{\jtt{keywords}} denote the reserved words in the language.
The over-lined $\overline{v}$ denotes a sequence  of syntactic entities $v$.
Both local and instance variables are denoted by $x$.
We assume distinguished local variables  $\jtt{this}$, $\jtt{myfuture}$, and  $\jtt{deadline}$ which
denote the actor itself, the  unique future corresponding to the process, and  its deadline,  respectively.
A distinguished instance variable $\jtt{time}$  denotes  the current time.
Any subscripted type $\mathsf{T}_{specialized}$ denotes a \textit{specialized} type of general type $\mathsf{T}$; 
e.g. $\mathsf{T}_{\jtt{Exception}}$ denotes all ``exception'' types.
A variable $f$ is in $\mathsf{T}_{future}$. 
\textsf{N} is a name (identifier) used for classes and method names.
%$e$ is in \textsf{Expression} ($E$);
$C$ denotes a class definition which consists of  a definition of its instance variables
and its methods;
$M_{sig}$ is a method signature; $M$ is a method definition; $S$ denotes a statement.
We abstract from the syntax the side-effect free expressions $e$ and
boolean expressions $b$.
%
%For simplicity, we drop the notation syntax to specify how $V$ can expand to different values along with arithmetic operations.
%\emph{Note} that small letters (e.g. $e$) are of type of their capital constructs (e.g. $E$).
}
\label{ch02:fig:grammar}
\end{figure}

@@

\parbf{What happens when a deadline is missed?}
Messages received by an actor generate processes.
Each actor contains one active process and all its other processes are queued. 
Newly generated processes are \emph{inserted} in the queue according to an application-specific policy.
When a \emph{queued} process misses its deadline it is \emph{removed} from the queue and a
corresponding \emph{exception} is recorded by its future (as described below).
When the currently active process is terminated the process at the head of the queue is activated
(and as such  dequeued).
The active process  cannot be \emph{preempted} and is forced
to  \emph{run to completion}.
In  Section \ref{ch02:sec:impl} we discuss the implementation details of this design choice.

@@

For example, in Listing \ref{ch02:lst:try:catch}, if the \gett \space operation raises an exception  control,  is transferred to line (5);
otherwise, the execution continues in line (3). 
In the \jtt{catch} block, the programmer has also access to the occurred exception that can be 
any kind of exception including an exception that is caused by a missed deadline.
In general, any uncaught exception gives rise to abortion of the active process and is recorded by its future.
Exceptions in our actor-based model thus are propagated by futures.

@@

% Since, we provide a reference implementation in Java (c.f. Section \ref{ch02:sec:impl}), we take advantage of exception ``stack traces''; 
% i.e. an instance of exception can produce a chain of ``causes'' that has led to the occurrence of the exception.
% This means that cathing an exception may in turn lead to throwing another exception or just rethrowing it. 
% Therefore, if the programmer decides not to use a \jtt{try-catch} block, the exception is propagated through the call hierarchy.
% It means that the propagation of the exception may also contribute to the stack trace of another exception.
% If any exception is not eventually caught, the execution environment will receive it and the application will stop.

% As discussed in detail in our previous work \cite{crisp-sac}, object $o'$ maintains a queue of processes for the incoming messages such as $m$.
% While a message $m$ is in the queue of $o'$, there is a chance that the deadline $d$ for message $m$ would be missed.
% We say a deadline $d$ for message $m$ on object $o'$ is missed when
% \begin{enumerate}
%  \item The process for message $m$ is \emph{not} the current process of $o'$. 
%  In other words, we do \emph{not} terminate the current process in case it missses its deadline; 
%  i.e. we allow \emph{run to completion} for the current process. And,
%  \item The response for message $m$ is not ready and the \emph{current} time has passed the deadline $d$ expectation.
% \end{enumerate}
% 
% When a deadline-miss is detected, it is required to be reflected in the future response that is being waited on in object $o$. 
% Therefore, for a deadline $d$ that is missed on message $m$:
% 
% \begin{enumerate}
%  \item The process that is related to message $m$ is aborted and removed from $o'$'s  process queue.
%  \item The aborted state is recorded in the corresponding future value $f$.
%  \item Whenever, \jtt{f.get()} is invoked, the \emph{aborted} state causes an exception. 
%  The exception is supposed to be handled by the user of the future; otherwise, it is propagated in the call hierarchy.
% \end{enumerate}
% 
% \emph{Starvation?} \Behrooz{Talk about the responsiveness of the system.}
% 
% \emph{Discussion?} Compare the approaches: event-based with no get() and exceptions as messages vs. messages with exceptions
\subsection{Case Study: Fredhopper Distributed Data Processing} \label{ch02:sec:frh}

\begin{figure}[h]
% \begin{wrapfigure}{r}{0.4\textwidth}
% \vspace{-40pt}
\begin{center}
\includegraphics[scale=0.4]{figs/DM}
\end{center}
% \vspace{-20pt}
\caption[Life cycle for remote data processing]{Fredhopper's Controller life cycle for remote data processing}
\label{ch02:fig:frh-dm}
% \vspace{-10pt}
% \end{wrapfigure}
\end{figure}

@@

To orchestrate different services such as FAS or data processing, Fredhopper takes advantage of a service controller (a.k.a. Controller).
Controller is responsible to passively manage different service installations for each customer.
For instance, in one scenario, a customer submits their data along with a processing request to their data hub server.
Controller, then picks up the data and initiates a data processing job (usually an ETL job) in a data processing service.
When the data processing is complete, the result is again published to customer environment and additionally becomes available through FAS services.
Figure \ref{ch02:fig:frh-dm} illustrates an example scenario that is described above.

@@

To illustrate the contribution of this paper, we extract a closed-world simplified version of the scenario in Figure \ref{ch02:fig:frh-dm} from Controller.
In Section \ref{ch02:sec:impl}, we provide an implementation-level usage of our work applied to this case study.

\section{Operational Semantics} \label{ch02:sec:sos}
We describe the semantics of the language by means of a two-tiered labeled transition system:
a local transition system describes the behavior of a single actor and a global transition system
describes the overall behavior of a system of interacting actors.
We define an actor state as a pair $\langle p,q \rangle$, where
\begin{itemize}
%  \item $o$ is the object identifier
 \item $p$ denotes the current active process of the actor, and
 \item $q$ denotes a queue of pending processes.
\end{itemize}
Each pending process is a pair  $(S, \tau)$ consisting of the current executing statement $S$ and the assignment $\tau$ of 
values  to the  \emph{local} variables (e.g., formal parameters).
The active process consists of a  pair  $(S,\sigma)$, where $\sigma$ assigns values to the local variables
and additionally assigns values to the instance variables
of the actor.

@@

\parbf{Incoming message.}
Let the active process $p$ belong to the actor $\tau(\jtt{this})$ (i.e., 
 $\sigma(\jtt{this})=\tau(\jtt{this})$ for the assignment $\sigma$ in $p$):
$$
\langle p , q, \phi\rangle \xrightarrow[]{m(\tau)}  \langle p , \mathsf{insert}(q, m(\overline{v}, d)), \phi\rangle
$$
where   $\mathsf{insert}(q, m(\tau))$  defines the result of inserting the process  $(S,\tau)$, where $S$ denotes the body of method $m$,
in $q$, according to some application-specific policy (described below in Section \ref{ch02:sec:impl}).

@@

% \subsection{Previously written in case notations would be reused}
% 
% \Behrooz{Previous wrong version. May be removed later.}
% \emph{Receive a new message.} The operational semantics when an object with identifier $o$ receives a message to invoke method $m$ with deadline $d$.
% 
% \begin{align}
% \frac
% {\mathsf{in} = m(\overline{v}, d) + \dot{p} = \mathsf{new()}}
% {(o, (\dot{p}.\jtt{deadline} \mapsto \mathsf{time()} + d,s[\overline{v}], d), p, \mathsf{enq}(q, \dot{p}), \sigma)}
% \end{align}
% 
% in which the operation \textsf{new()} creates a new instance of a process for the concurrent object.
% Additionally, as we discssued in Equation \ref{ch02:eq:orderedq}, each concurrent object owns an ordered queue for the processes.
% The process queue is ordered based on the result of $\sigma$ function:
% 
% $$ \sigma(\dot{p}) = \dot{p}.\jtt{deadline} $$
% 
% Thus, the operational semantics of \textsf{enq} ensures to find a proper index place $i$ for the new process $\dot{p}$ such that:
% 
% $$
% \forall k < i \::\: q_k.\jtt{deadline} < \dot{p}.\jtt{deadline} \wedge \forall k > i \::\: q_k.\jtt{deadline} > \dot{p}.\jtt{deadline}
% $$
% 
% And, then, $q_i = \dot{p}$ completes the enqueuing action. Since, $q$ is an \emph{ascending} ordered queue, the required action to deqeue the 
% ``next'' process is simply to remove and return the \emph{head} of the queue; namely $q_0$.
% 
% \emph{Deadline Miss Detection.} A deadline miss is detected by a separate watchdog thread of the active concurrent object.
% The watchdog thread observes all the processes but the current process of the object to allow it complete its execution:
% 
% \begin{align}
% \frac
% {\forall p \neq \overline{p} \::\: (o,s[f_p \mapsto \jtt{null}],\overline{p},q,\sigma) + p.\jtt{deadline} < \mathsf{time()}}
% {(o,s[f_p.\jtt{aborted} \mapsto \jtt{true}], \overline{p}, q - p, \sigma)}
% \end{align}
% 
% Consequently, when the future of a missed-deadline process $p$ is accessed, it causes the application to generate and throw an exception:
% 
% \begin{align}
% \frac
% {(o,(f.\jtt{get()},s[f.\jtt{aborted} = \jtt{true}],d),q,\sigma)}
% {(o,(x \mapsto \jtt{new Exception},s[f \mapsto \jtt{null}],d),q,\sigma)}
% \end{align}

\section{Implementation} \label{ch02:sec:impl}
We base our implementation on Java's concurrent package: \jtt{java.util.concurrent}. 
The implementation consists of the following major components:
\begin{enumerate}
 \item An extensible language API that owns the core abstractions, architecture, and implementation. 
For instance, the programmer may extend the concept of a scheduler to take full control of how, i.e., in what order, the processes of the individual actors are queued
(and as such scheduled for execution).
We illustrate the scheduler extensibility with an example in the case study below.
 \item Language Compiler that translates the modeling-level programs into Java source.
 We use ANTLR \cite{antlr} parser generator framework to compile modeling-level programs to actual implementation-level source code of Java.
 \item The language is seamlessly integrated with Java.
 At the time of programming, language abstractions such as data types and third-party libraries from either Crisp or Java are equally usable by the programmer.
\end{enumerate}

@@

\parbf{Fredhopper case study.}
As introduced in Section \ref{ch02:sec:frh}, we extract a closed-world simplified version from Fredhopper Controller.
We apply the approach discussed in this paper to use deadlines for asynchronous messages.
% Controller operates in a high-demand distributed setting as in Figure \ref{ch02:fig:ctl:deploy}.

Listing \ref{ch02:lst:frh:polling} and \ref{ch02:lst:frh:deadlines} present the difference in the previous Controller and the approach in Crisp.
The left code snippet shows the Controller that uses polling to retrieve data processing results.
The right code snippet shows the one that uses messages with deadlines.

@@

When the approach in Crisp in the right snippet is applied to Controller, it is guaranteed that all data job requests are terminated in a \emph{finite} amount of time.
Therefore, there cannot be complains about never receiving a response for a specific data job request.
Many of Fredhopper's customers rely on data jobs to eventually deliver an e-commerce service to their end users.
Thus, to provide a guarantee to them that their job result is always published to their environment is critical to them.
As shown in the code snippet, if the data job request is failed or aborted based on a deadline miss, 
the customer is still eventually informed about the situation and may further decide about it.
However, in the previous version, the customer may never be able to react to a data job request because its results are never published.
% \begin{wrapfigure}{r}{0.35\textwidth}
% \vspace{-35pt}
% \begin{center}
% \includegraphics[scale=0.35]{frh-ctl}
% \end{center}
% \vspace{-20pt}
% \caption{Fredhopper's Controller distribution model}
% \label{ch02:fig:ctl:deploy}
% \vspace{-15pt}
% \end{wrapfigure}

In comparison to the Controller using polling, there is a way to express timeouts for future values.
However, it does not provide language constructs to specify a deadline for a message that is sent to data processing service.
A deadline may be simulated using a combination of timeout and periodic polling approaches (Listing \ref{ch02:lst:frh:polling}).
Though, this approach cannot guarantee eventual termination in all cases; as discussed before that Step 4 in Figure \ref{ch02:fig:frh-dm} may never complete.
Controller is required to meet certain customer expectations based on an SLA.
Thus, Controller needs to take advantage of a language/library solution that can provide a higher level of abstraction for real-time scheduling of concurrent messages.
When messages in Crisp carry a deadline specification, Controller is able to guarantee that it can provide a response to the customer.
This termination guarantee is crucial to the business of the customer.

@@

In the above listings, Listing \ref{ch02:lst:frh:sched} defines a custom scheduler that determines the priority of two processes with custom logic for specific customer.
To use the custom scheduler, the only requirement is that the class \jtt{DataProcessor} defines a specific class variable called \jtt{scheduler} in Listing \ref{ch02:lst:frh:dm}.
The \textsl{custom scheduler} is picked up by Crisp core architecture and is used to schedule the queued processes.
% In the listing, \jtt{DefaultSchedulingManager} denotes the default EDF scheduler that is implemented in Crisp.
Thus, all processes from customer \jtt{A} have priority over processes from other customers no matter what their deadlines are.

@@

\begin{table}[h]
% \begin{wraptable}{l}{0.25\textwidth}
% \vspace{-15pt}
\begin{tabular}{c c}\thickhline
$s_1$ & $s_2$ \\ \thickhline
$88.71\%$ & $94.57\%$ \\ \hline
\end{tabular}
\caption{Evaluation Results}
\label{ch02:tbl:eval}
% \vspace{-15pt}
% \end{wraptable}
\end{table}

For a customer $c$, the satisfaction can be denoted by $s = \frac{r^F_c}{r_c}$; in which $r^F_c$ is the number of finished data processing jobs and $r_c$ is the total number of requested data processing jobs from customer $c$. 
We extracted statistics for completed and never-ended data processing jobs from Controller logs ($s_1$).
We replayed the logs with Crisp approach and measured the same property ($s_2$).
We measured the same property for 180 customers that Fredhopper manages on the cloud.
In this evaluation, a total number of about 25000 data processing requests were included.
The results show $6\%$ improvement in Table \ref{ch02:tbl:eval} (that amounts to around 1600 better data processing requests). 
Because of data issues or wrong parameters in the data processing requests, there are requests that still fail or never end and should be handled by a human resource.

@@

\section{Related Work} \label{ch02:sec:relwork}
The programming language presented in this paper is a real-time  extension of the  language introduced in  \cite{crisp-sac}. This new extension features
\begin{itemize}
 \item integration of  asynchronous messages with  deadlines and futures with timeouts;
 \item  a general mechanism for handling exceptions raised by missed deadlines;
  \item high-level specification of application-level scheduling policies; and
  \item a formal operational semantics.
\end{itemize}
To the best of our knowledge the resulting language is the first implemented 
real-time actor-based programming language which formally integrates the above features.
